---
title: "SGUR map"
author: "FC"
date: "2024-07-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
library(ggplot2)
library(tidyverse)
```

#Get Data

I downloaded the Urban Rural shape file (and others that came with it) from https://spatialdata.gov.scot/geonetwork/srv/api/records/564db46c-3153-423c-ac84-90d41ec2652c on 16/07/24

I downloaded the Health Board shape file (and others that came with it) from https://spatialdata.gov.scot/geonetwork/srv/api/records/f12c3826-4b4b-40e6-bf4f-77b9ed01dc14 on 16/07/24

I downloaded the Hospitals shape file (and others that came with it) from https://data.spatialhub.scot/dataset/nhs_hospitals-is on 16/07/24

```{r}
UR_map <- read_sf("data/UR/SG_UrbanRural_2020.shp")

HB_map <- read_sf("data/HB/SG_NHS_HealthBoards_2019.shp")

hospitals_map <- read_sf("data/Hospitals/pub_hosp.shp")
```

#View separate maps

##UR Map

First, change the 2-fold category names
```{r}
UR_map <- UR_map %>%
  mutate(UR2Name = case_when(
    UR2Name == "Rural Areas" ~ "Rural",
    UR2Name == "Rest of Scotland" ~ "Urban",
    TRUE ~ UR2Name
    ))
```



```{r}
# UR_map %>%
#   ggplot(data = UR_map) +
#   geom_sf(aes(fill = UR2Name), size = 0.01) + 
#   theme_minimal() +  # Use theme_minimal for a simple clean theme
#   labs(title = "Urban and Rural Areas",
#        fill = "Classification")
```

##Health Board Map

###All
```{r}
HB_map %>%
  ggplot() +
  geom_sf(aes(fill = HBName), size = 0.01) +
  theme_void()
```


##Hospitals Map
These are just data points, you won't see the shape of anything under them. 

###Plotting
```{r}
hospitals_map %>%
  ggplot() +
  geom_sf(data = hospitals_map, color = "red", size = 1) + # this size is for the markers
  geom_text(data = hospitals_map, aes(label = hosp_name, geometry = geometry), 
            stat = "sf_coordinates", size = 3, nudge_y = 0.01, nudge_x = 0.01, check_overlap = FALSE) + # check_overlap = TRUE would exclude any overlapping labels
  theme_void()
```


#Join maps

Filter the HB map to just Grampian
```{r}
Grampian_map <- HB_map %>% filter(HBName == "Grampian")

```

Join the Grampian map to the UR map
```{r}
UR_Grampian_map <- UR_map %>%
  st_intersection(Grampian_map) %>%
  st_join(Grampian_map)
```

Create the list of IC hospitals
```{r}
IC_hosps <- c("Aboyne Hospital", "Chalmers Hospital", "Fraserburgh Hospital", "Glen O'Dee Community Hospital",  "Insch War Memorial Hospital", "Inverurie Hospital", "Jubilee Hospital", "Kincardine Community Hospital", "Peterhead Community Hospital", "Turriff Cottage Hospital", 
              "Ugie Hospital", "Fleming Cottage Hospital", "Seafield Hospital", "Stephen Cottage Hospital", "Turner Memorial Hospital") 
IC_map <- hospitals_map %>%
  filter(hosp_name %in% IC_hosps) %>%
  st_intersection(Grampian_map)
```


# Plot Grampian UR IC map

copying Frank's code for the intcare theme
```{r}
theme_intcare <- function() {theme_set(theme_classic()) +
    theme(legend.title = element_blank(),
          legend.position = "top",
          legend.text = element_text(size=14),
          axis.text = element_text(size=11),
          axis.title.y = element_text(size=14))}
```

plotting

the UR2Name fill defaulted to swapped colours. My other figures have the blue-ish for rural and pink-ish for urban, so this needs to match. I couldn't find the exact colour codes but I think I've got close enough. 

This isn't perfect; the right most label is cut off. I tried various genAI advice but nothing worked. 
```{r}
UR_IC_grampian_map <- 
  ggplot() +
  geom_sf(data = UR_Grampian_map, aes(fill = UR2Name), color = NA) + # color = NA removes odd squiggles all across Grampian
  geom_sf(data = IC_map, color = "red", size = 2, shape = 17) + # this is the fill for the hospital markers
  scale_fill_manual(values = c("Rural" = "#00BFC4",   # these aren't the exact colour codes, I had to eyeball it. 
                               "Urban" = "#FF8080")) + 
  geom_text(data = IC_map, aes(label = hosp_name, geometry = geometry), size = 3, nudge_y = 0.01, nudge_x = -5, check_overlap = FALSE, stat = "sf_coordinates", clip = FALSE) +
  labs(title = "Grampian Health Board Area Map",
       fill = "Urban/Rural Classification") +
  theme_intcare()

plot(UR_IC_grampian_map)

ggsave("outputs/UR_IC_grampian_map.png", plot = UR_IC_grampian_map, width = 10, height = 8, dpi = 300)
```

